Make any stupid plot you want

### 2021-02-25 draft/sections/math.tex
tweaking of monoid/fiber language

As discussed in the introduction, visualization is generally defined as structure preserving maps from data to graphic representation. In order to formalize this statement, we describe the connectivity of the records using topology and define the structure on the components in terms of the monoid actions on the component types. By formalizing structure in this way, we can evaluate the extent to which a visualization preserves the structure of the data it is representing and build structure preserving visualization tools. 
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}. Monoid actions are a set of operations and are discussed in detail in section~\ref{sec:data_monoid}. Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are symmetric 
\begin{equation}
\mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
\end{equation}
such that applying a monoid $m \in \monoid$ transformation to the input $\delement \in \mathscr{\dtotal}$ to $\mathscr{\vartist}$ is equivalent to applying a monoid $\varphi(\monoid) \in \monoid^{\prime}$ to the output of the artist $\vartist(\delement) \in \mathscr{\gtotal}$.

@@

For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree K temperature measurements taken at those times. Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$. Temperature values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$. The fiber is 
\begin{equation}
    \ftotal =  \reals \times \realsp 
\end{equation} 
where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\realsp)$ and is the set of values $\fttype=\realsp$. In figure~\ref{fig:fiber_example_cube}, temperature is replaced with wind. This wind variable is of type \texttt{wind} and has two components speed and direction $\{(s,d) \in \reals^{2} \mid  0\leq s,\, 0 \leq d \leq 360\}$ . Therefore, the fiber is 
\begin{equation}
    \dfiber = \realsp \times \reals^{2}
\end{equation} 
such that $F_1$ is specified by $(\fname = wind,\, \ftype = \texttt{wind},\, \fttype=\reals^{2})$.  As illustrated in figure~\ref{fig:data_fiber_example}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.

@@

\begin{table}[H]
    \renewcommand{\arraystretch}{2}
    \begin{tabulary}{\textwidth}{|l|L|l|}\hline
     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i})}$  \\ \hline                                              
    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
    \end{tabulary}
    \caption{Some possible components of the fiber \vfiber\ for a visualization function implemented in Matplotlib}
    \label{tab:mpl_visual_variable_fiber}
\end{table}

@@

This equivariance constraint yields guidance on what makes for an invalid transform. For example, given a visual fiber of the same type as the data fiber $\dfiber=\vfiber$, the transform $\vchannel_{i}(x) = .5$ does not commute under translation monoid action $t(x) = x+2$  
\begin{align}
    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + \vchannel(2)\\
    .5 &\neq .5 + .5
\end{align}

On the other hand figure~\ref{fig:artist_nu} illustrates a valid \vchannel\ mapping from \textbf{Strings} to symbols. The group action on these sets is permutation, so shuffling the words must have an equivalent shuffle of the symbols they are mapped to. Continuing with the assumption that $\dfiber=\vfiber$, we can describe the constraints on the other Steven's measurment group types. To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$ 
\begin{equation}
\text{ if } delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
\end{equation}
the visual encodings must also have some sort of ordering. For interval scale data, \vchannel\ is equivariant under translation monoid actions if 
\begin{equation}
\vchannel (x + c) = \vchannel(x) + \vchannel(c)
\end{equation}
while for ratio data, there must be equivalent scaling 
\begin{equation}
\vchannel(xc) = \vchannel(x)\vchannel(c) 
\end{equation}

The assumption $\dfiber=\vfiber$ is made here so that we could omit  $\varphi: \monoid\rightarrow\monoid^{\prime}$ maps that convert the monoid action on \dfiber\ to the equivalent action on \vfiber. As shown in table~\ref{tab:mpl_visual_variable_fiber} though, it is not uncommon for data and visual variables to resolve to the same types. The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions are equivariant; they also provide guidance on constructing new equivariant \vchannel\ functions.

@@

While \gsection\ generates the entire graphic, we will restrict the discussion of \vmark\ to generation of sections of a glyph. We formally describe a glyph as \vmark\ applied to the regions \dbasepoint that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
\begin{equation}
\dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
\end{equation}
where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1]. We define the glyph as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
\begin{equation}
    \begin{tikzcd}
        \gtotal \arrow[r, shift left] & \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] &  & \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
        \end{tikzcd}
    \label{eq:mark}
\end{equation}
such that for every glyph there is at least one corresponding section on \dbase. This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}. The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.

@@

The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back $\vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber$ copies the visual fiber \vtotal over the preimage in \vpreimg.

This is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase.  All the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\ are in the preimage \vpreimg.  Given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map from \vsection on a point \dbasepoint, there is a direct map from the visual section over data base space  $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ to the visual section \vsectionpull\ over graphic base space. This map means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint. This means that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$, which is illustrated in figure~\ref{fig:artist_q_hat} as each dot on \vfiber is equivalent to the line intersection \vfiberpull.

Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out. When \vmark\ maps visual sections  into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to the pulled back visual section \vsectionpull\ then 
\begin{equation}
    $\gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)$
\end{equation}
the graphic section \rho\ evaluated on a visual region \gbasepoint\ is defined as the assembly function \vmark\ with input pulled back visual section \vsectionpull\ also evaluated on \gbasepoint. Since the pulled back visual section \vsectionpull\ is the visual section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function 
\begin{equation}
\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
\end{equation} 
where the assembly function \vmarkd\ that takes as input the visual section on data \vsection\ is defined to be the assembly function \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gspace.

>>>>

As discussed in the introduction, visualization is generally defined as structure preserving maps from data to graphic representation. In order to formalize this statement, we describe the connectivity of the records using topology and define the structure on the components in terms of the monoid actions on the component types. By formalizing structure in this way, we can evaluate the extent to which a visualization preserves the structure of the data it is representing and build structure preserving visualization tools. 
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in section~\ref{sec:data_monoid}. Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are symmetric 
\begin{equation}
\mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
\end{equation}
such that applying a monoid action $m \in \monoid$ to the input $\delement \in \mathscr{\dtotal}$ to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the output of the artist $\vartist(\delement) \in \mathscr{\gtotal}$.

@@

For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree^K temperature measurements taken at those times. Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$. Temperature values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$. The fiber is 
\begin{equation}
    \ftotal =  \reals \times \realsp 
\end{equation} 
where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\realsp)$ and is the set of values $\fttype=\realsp$. In figure~\ref{fig:fiber_example_cube}, temperature is replaced with wind. This wind variable is of type \texttt{wind} and has two components speed and direction $\{(s,d) \in \reals^{2} \mid  0\leq s,\, 0 \leq d \leq 360\}$ . Therefore, the fiber is 
\begin{equation}
    \dfiber = \realsp \times \reals^{2}
\end{equation} 
such that $F_1$ is specified by $(\fname = wind,\, \ftype = \texttt{wind},\, \fttype=\reals^{2})$.  As illustrated in figure~\ref{fig:data_fiber_example}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.

@@

\begin{table}[H]
    \renewcommand{\arraystretch}{2}
    \begin{tabulary}{\textwidth}{|l|L|l|}\hline
     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}$  \\ \hline                                              
    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
    \end{tabulary}
    \caption{Some possible components of the fiber \vfiber\ for a visualization function implemented in Matplotlib}
    \label{tab:mpl_visual_variable_fiber}
\end{table}

@@

This equivariance constraint yields guidance on what makes for an invalid transform. For example, given a visual fiber of the same type as the data fiber $\dfiber_i=\vfiber_i$, the transform $\vchannel_{i}(x) = .5$ does not commute under translation $t(x) = x+2$  
\begin{align}
    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + \vchannel(2)\\
    .5 &\neq .5 + .5
\end{align}

On the other hand figure~\ref{fig:artist_nu} illustrates a valid \vchannel\ mapping from \textbf{Strings} to symbols. The group action on these sets is permutation, so shuffling the words must have an equivalent shuffle of the symbols they are mapped to. Continuing with the assumption that $\dfiber=\vfiber$, we can describe the constraints on the other Steven's measurment group types. To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$ 
\begin{equation}
\text{ if } delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
\end{equation}
the visual encodings must also have some sort of ordering. For interval scale data, \vchannel\ is equivariant under translation monoid actions if 
\begin{equation}
\vchannel (x + c) = \vchannel(x) + \vchannel(c)
\end{equation}
while for ratio data, there must be equivalent scaling 
\begin{equation}
\vchannel(xc) = \vchannel(x)\vchannel(c) 
\end{equation}
The assumption $\dfiber_i=\vfiber_i$ is made here so that we could omit  $\varphi: \monoid\rightarrow\monoid^{\prime}$ maps that convert the monoid action on $\dfiber_{i}$ to the equivalent action on $\vfiber_{i}$. As shown in table~\ref{tab:mpl_visual_variable_fiber}, it is not uncommon for data and visual variables to resolve to the same types. The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions are equivariant and also provide guidance on constructing new equivariant \vchannel\ functions.

@@

While \gsection\ generates the entire graphic, we will restrict the discussion of \vmark\ to generation of sections of a glyph. We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
\begin{equation}
\dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
\end{equation}
where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1]. We define the glyph as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
\begin{equation}
    \begin{tikzcd}
        \gtotal \arrow[r, shift left] & \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] &  & \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
        \end{tikzcd}
    \label{eq:mark}
\end{equation}
such that for every glyph there is at least one corresponding section on \dbase. This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}. The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.

@@

The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\
\begin{equation}
    \vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber
\end{equation}
copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\. This set of points \gbasepoint\ are the preimage \vpreimg\ of \dbasepoint.

This copying is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase. Given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map from \vsection on a point \dbasepoint, there is a direct map from the visual section over data base space  $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ to the visual section \vsectionpull\ over graphic base space. This map means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint. This means that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$, which is illustrated in figure~\ref{fig:artist_q_hat} as each dot on \vfiber is equivalent to the line intersection \vfiberpull.

Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out. When \vmark\ maps visual sections  into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to the pulled back visual section \vsectionpull\ then 
\begin{equation}
    \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
\end{equation}
the graphic section \gsection\ evaluated on a visual region \gbasepoint\ is defined as the assembly function \vmark\ with input pulled back visual section \vsectionpull\ also evaluated on \gbasepoint. Since the pulled back visual section \vsectionpull\ is the visual section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function 
\begin{equation}
\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
\end{equation} 
where the assembly function \vmarkd\ that takes as input the visual section on data \vsection\ is defined to be the assembly function \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gspace.
